---
title: "hw2"
format: html
editor: 
  markdown: 
    wrap: 72
---

# Background

Present-day environmental justice may reflect legacies of injustice in
the past. The United States has a long history of racial segregation
which is still visible. During the 1930’s the Home Owners’ Loan
Corporation (HOLC), as part of the New Deal, rated neighborhoods based
on their perceived safety for real estate investment. Their ranking
system, (A (green), B (blue), C (yellow), D (red)) was then used to
block access to loans for home ownership. Colloquially known as
“redlining”, this practice has had widely-documented consequences not
only for community wealth, but also health.1 Redlined neighborhoods have
less greenery2 and are hotter than other neighborhoods.3

Check out coverage by the New York Times.

A recent study found that redlining has not only affected the
environments communities are exposed to, it has also shaped our
observations of biodiversity.4 Community or citizen science, whereby
individuals share observations of species, is generating an enormous
volume of data. Ellis-Soto and co-authors found that redlined
neighborhoods remain the most undersampled areas across 195 US cities.
This gap is highly concerning, because conservation decisions are made
based on these data.

```{r}
library(tidyverse)
library(here)
library(sf) 
library(tmap)
library(spData)
```

```{r}
ejscreen <- st_read(here("data", "ejscreen",
                        "EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb")) 
birds <- st_read(here("data", "gbif-birds-LA", "gbif-birds-LA.shp"))
holc <- st_read(here("data", "mapping-inequality", "mapping-inequality-los-angeles.json"))
```

# Part 1: Legacy of redlining in current environmental (in)justice

## Description

1.  For this assignment, you must complete the following:

Create a map of historical redlining neighborhoods, including:
neighborhoods colored by HOLC grade an appropriate base map a clear
title and legend

```{r}
world <- spData::world
la <- ejscreen %>% 
  filter(CNTY_NAME == 'Los Angeles County')

states <- spData::us_states
```

```{r}
# Checking to see if they have the same CRS
st_crs(la) == st_crs(holc)
```
```{r}
# False- Need to set them equal
la_transform <- st_transform(la, crs = st_crs(holc))

st_crs(la_transform) == st_crs(holc)
```

```{r}
tm_shape(la_transform, 
         bbox= c(-118.65, 33.7, -117.7, 34.3)) +
  tm_polygons() +
  tm_graticules() +
tm_shape(holc) +
  tm_polygons(fill = 'grade',
              fill.scale = tm_scale(values = c("forestgreen", 
                                               "navy",
                                               "gold1",
                                               "red4"))) 
  #tm_shape() +
  #tm_polygons()
```
```{r}

```


## Creating a summary table
```{r}
holc_la <- st_join(x = la_transform, y = holc,
                   join = st_overlaps) %>%
  st_drop_geometry()
```


2.  Create a table summarizing: the percentage of census block groups
that fall within each HOLC grade Also include the percent of census
    block groups that don’t fall within a HOLC grade Hint: The HOLC data
    contains the grades and the EJScreen data contains the census
    blocks, so you will need to combine the data spatially before doing
    summary statistics. Once you combine and no longer need the
    geometries, you can use st_drop_geometry()

```{r}
holc_la_summary <- holc_la %>% 
  group_by(grade) %>% 
  summarise(pop_sum = sum(ACSTOTPOP)) %>% 
  ungroup() %>% 
  filter(grade == NA)
```


```{r}
holc_la$grade
```


3.  Create at least two visualizations summarizing current conditions
    (from the EJScreen data) within HOLC grades using the mean of the
    following variables (you may combine variables or create separate
    plots): % low income percentile for Particulate Matter 2.5
    percentile for low life expectancy Use ggplot for your
    visualizations! You will first need to calculate mean of each
    variable grouped by HOLC grade.

4.  Write a brief paragraph reflecting on these results Interpret the
    patterns you observe in your results Discuss potential relationships
    between historical redlining grades and current
    environmental/socioeconomic conditions

# Part 2: Legacy of redlining in biodiversity observations

## Description
For this assignment, you must produce the following based on observations from 2021-2023:

A figure summarizing the percent of observations within redlined neighborhoods within each HOLC grade

Create a visualizations that shows:

The percentage of bird observations within each HOLC grade
Include an appropiate title, axis labels, and legend
Hints: Ensure the bird observations and HOLC dataset have matching CRS’, then perform a spatial join to assign each bird observations to a corresponding HOLC grade.
Spolier alert!! Our results don’t match the findings from Ellis-Soto et al. 2023! Read the abstract of the study. Why might we have obtained different results in our analysis? What did the paper consider that we did not?
